% A ConTeXt document [master document: texts.tex]

\section[title=Strings]

\starttyping
// files.c
#include <stdlib.h>
#include <string.h>
#include <assert.h>

#include "joyLoL/macros.h"
#include "joyLoL/coAlg/coAlgs.h"
#include "joyLoL/lists.h"
#include "joyLoL/text.h"
#include "joyLoL/text_private.h"

// Texts are a collection of characters, which are used by the Parser
// to extract successive symbols.
//
// Texts are created on one of three backing suppliers of characters:
// 1. a single string
// 2. a NULL terminated array of strings
// 3. an external file
// 4. a readline interaction with a user
//
// In all four cases, the Parser's nextSymbol method requests successive
// **lines** of characters (deliminated by new-line-characters).
//
// It is critical, for correct interaction with the user via readline,
// that the initial line is NOT obtained until actually requested by
// the parser's nextSymbol method.
//
// It is also critical that once completed, none of the sources, get
// asked for subsequent lines.
//
// When the text has been completed, the nextLine function ensures
// that aText->curLine is NULL.

/////////////////////////////////////////////////////////////////////////////
// Text from file
//

void nextLineFromFile(Text* aText) {
  DEBUG(FALSE, "->nextLineFromFile %s\n", "");
  if (!aText) return; // there is nothing we can do!
  if (!aText->inputFile) return; // there is nothing we can do!

  // getline returns alloc'ed memory so we need to free it here.
  if (aText->curLine) free((void*)aText->curLine);

  aText->curLine    = NULL;
  aText->curChar    = NULL;
  aText->lastChar   = NULL;

  char *linePtr    = NULL;
  size_t n         = 0;
  if (!getline(&linePtr, &n, aText->inputFile) && n < 1) {
    DEBUG(FALSE, "<-nextLineFromFile %s\n", "getline returned 0");
    aText->completed = TRUE;
    return;
  }

  if (feof(aText->inputFile)) {
    // we have already reached the end of the text;
    DEBUG(FALSE, "<-nextLineFromFile %s\n", "feof returned file end");
    aText->completed = TRUE;
    aText->curLine   = NULL;
    aText->curChar   = NULL;
    aText->lastChar  = NULL;
    return;
  }

  aText->curLine    = linePtr;

  aText->curChar    = aText->curLine;
  aText->lastChar   = aText->curChar + strlen(aText->curLine);
  DEBUG(FALSE, "<-nextLineFromFile [%s]\n", aText->curLine);
}

Text* createTextFromInputFile(CoAlgebras* coAlgs,
                              FILE *anInputFile) {
  Text* aText = (Text*)calloc(1, sizeof(Text));
  assert(aText);
  //
  // external file specific initializations
  //
  aText->inputFile  = anInputFile;
  aText->nextLine   = nextLineFromFile;
  //
  // general initializations
  //
  assert(coAlgs);
  aText->coAlgebras = coAlgs;
  aText->completed  = FALSE;
  aText->sym        = NULL;
  aText->curLine    = NULL;
  aText->curChar    = NULL;
  aText->lastChar   = NULL;
  //
  aText->textLines  = NULL;
  aText->curLineNum = 0;
  //
  aText->newPrompt      = NULL;
  aText->continuePrompt = NULL;
  aText->curPrompt      = NULL;
  return aText;
}
\stoptyping
