%% Time 

\chapter{Time: Graphs, paths and refinements}

In anticipation of our future work, we take a slightly non-standard route towards defining
(trans-finite) computation. We leave it to the reader to understand both why we take
this route, and that we have, in fact, provided the same structures.

The classical theory of computation, with its (almost exclusive) focus on
\emph{terminating} computational processes, is largely focused on \emph{time} to the
exclusion of \emph{space}. For \emph{terminating} processes, any parallel (spacial)
process can be serialized. This serialisability \emph{is} equivalent to set theory's
Well-ordering Theorem and hence the Axiom of Choice. As we will see below,
\emph{non-terminating} processes need not be globally serializable. This is, again,
equivalent to the lack of global time in relativistic space-time.

Since we are not omni-potent beings, while we can prove theorems about non-terminating 
processes, we can not directly experience them. We can only access non-terminating processes 
via (terminating) ordinal approximations to the full non-terminating process. This means 
that, in a sense, ``all'' computation can \emph{effectively} be reduced to serializable 
terminating processes. 

In this section we explore the \emph{time} focused foundations of computations. In the next 
section we explore the addition of \emph{space-time} (and hence parallelism).

One of the central themes of this cycle of papers is the importance of (combinatorial)
graphs, paths and (co-)algebraic refinements. We start by defining (trans-finite) graphs.

There are essentially two distinct ways to define a mathematical structure: logically, by
explicitly assuming the existence of a set with specific (logically expressed) properties,
or computationally, in categorical language, as the fixed point of an endofunctor. From a
non-categorical inductive point of view, an early expression of this dichotomy can be
found in \cite[page 1]{moschovakis1974induction}. In this paper we follow the logical
approach. In subsequent papers in this cycle, we will use the computational approach.

There are relatively few discussions of non-finite graphs from which we mention Reinhard
Diestel's work, \cite{diestel2006graphTheory, diestel1990infiniteGraphTheory}.
Unfortunately, for trans-finite graphs, we must go, more or less, ``on beyond
zebra''\footnote{Growing up, one of my favourite books was Dr Seuss' ``On beyond Zebra'',
\cite{seuss1955onBeyondZebra}. \emph{Be warned}: One of my greatest faults is that I seem
congenitally incapable of staying within ``obvious'' bounds...}\footnote{Another of my
greatest faults is that my mind is a web of snippets of ideas. Linearizing this web into
single narrative, is \emph{very} hard for me to do, and may be hard to fully follow until 
you have heard ``the whole shaggy-dog story''.}.

Our definition of graph is inspired by the definition of metagraph in
\cite{macLane1971categoriesWorkingMathematician} section I.1. Note that our definition is
a generalisation of most typical definitions, for example \cite[section
1.1]{diestel2006graphTheory}. In particular, all of our graphs are \emph{explicilty}
\emph{directed}. Categorically, a graph is simply a presheaf on the category
$ \cdot \substack{\mathbf{\longrightarrow} \\[-0.7ex] \mathbf{\longrightarrow}} \cdot $

\begin{definition}
A \define{graph}{}, $G(V, E, \pi_0, \pi_1)$, is a set, $V$, of \define{verticies}{}, and a
set, $E$, of \define{edges}{}, together with two maps, \map{\pi_0}{E}{V} and
\map{\pi_1}{E}{V}, called the source and target maps.

A \define{labeled graph}{}, $lG(V, E, L_V, L_E, \pi_0, \pi_1, l_V, l_E)$, is a graph $G(V,
E, \pi_0, \pi_1)$ together with a set, $L_V$, of \define{vertex labels}{}, and a set,
$L_E$, of \define{edge labels}{}, and maps, \map{l_V}{V}{L_V}, and \map{l_E}{E}{L_E}.
Often the set of vertex labels, $L_V$, and vertex label map, $l_V$, are omitted (or
alternatively the set of vertex labels is assumed to be the trivial one point set).
 
Given a labeled graph, $lG(V, E, L_V, L_E, \pi_0, \pi_1, l_v, lE)$ its \define{underlying
graph}{}, is the graph, $G(V, E, \pi_0, \pi_1)$.
\end{definition}

\begin{definition}
A \define{graph morphism}{}, \map{\phi}{G_0}{G_1}, is a map between the graphs,
$G_0(V^{G_0}, E^{G_0}, \pi^{G_0}_0, \pi^{G_0}_1)$ and $G_1(V^{G_1}, E^{G_1}, \pi^{G_1}_0,
\pi^{G_1}_0)$ for which
\begin{enumerate}
\item $\phi(V^{G_0}) \subset V^{G_1}$
\item $\phi(E^{G_0}) \subset E^{G_1}$
\item $\phi \compose \pi^{G_0}_i = \pi^{G_1}_i \compose \phi$ for $i \in {0, 1}$
\end{enumerate}

A \define{labeled graph morphism}{}, \map{\phi}{lG_0}{lG_1}, between a pair of labeled
graphs, $lG_0(V^{lG_0}, E^{lG_0}, L^{lG_0}_V, L^{lG_0}_E, \pi^{lG_0}_0, \pi^{lG_0}_1,
l^{lG_0}_V, l^{lG_0}_V)$ and $lG_1(V^{lG_1}, E^{lG_1}, L^{lG_1}_V, L^{lG_1}_E,
\pi^{lG_1}_0, \pi^{lG_1}_1, l^{lG_1}_V, l^{lG_1}_V)$, is a graph morphism between the
underlying graphs, $G_0$ and $G_1$ for which
\begin{enumerate}
\item $\phi(L^{lG_0}_V) \subset L^{lG_1}_V$
\item $\phi(L^{lG_0}_E) \subset L^{lG_1}_E$
\item $\phi \compose l^{lG_0}_V = l^{lG_1}_V \compose \phi$
\item $\phi \compose l^{lG_0}_E = l^{lG_1}_E \compose \phi$
\end{enumerate}
\end{definition}

\begin{definition} \TODO{expand this to include isomorphism, monomorphism, epimorphisms}
A \define{graph embedding}{} is a graph morphism, \map{\phi}{G_0}{G_1}, respectively, a
labelled graph morphism, \map{\phi}{lG_0}{lG_1}, which is, categorically, a monomorphism.
\end{definition}

\begin{lemma}
Given a graph, $G(V, E, \pi_0, \pi_1)$, there exists a \define{dual graph}{},
$\tilde{G}(\tilde{V}, \tilde{E}, \tilde{\pi}_0, \tilde{\pi}_1)$, unique up to graph 
isomorphism, for which
\begin{enumerate}
\item $\tilde{V} = V$,
\item $\tilde{E} = \set{(v, u) \suchThat (u, v) \in E}$
\end{enumerate}
\end{lemma}

\begin{proof}
\TODO{PROVE THIS!}
\end{proof}

\TODO{Do we need the equivilant lemma for dual labeled graphs?}

Our next goal is to provide a definition of $1$-dimensional (time) ``$1$-paths''.
Intuitively a (directed) ``$1$-path'' \emph{in} a graph is an embedding of a ``totally
ordered'' directed graph.

\TODO{NOTE: the current use of $(x,y) \in E$ is not general enough for our purposes. It 
assumes that there is a unique edge between the vertices $x$ and $y$. For paths this is ok, 
for general graphs it is NOT. We need to rework this notation.}

\begin{definition}
A \define{(directed) 1-path}{}, $P(V, E, \pi_0, \pi_1)$, is a graph for which:

\begin{enumerate}
\item \define{Anti-symmetry:}{} for any pair of vertices, $x, y \in V$, if $(x,y) \in E$ and
$(y,x) \in E$ then $x = y$.
\item \define{Transitivity:}{} for any triple of verticies, $x, y, z \in V$, 
if $(x,y) \in E$ and $(y,z) \in E$ then $(x,z) \in E$.
\item \define{Total order:}{} for any pair of vertices, $x, y \in V$, if $x \neq y$ then
either $(x,y) \in E$ or $(y,x) \in E$ but not both.
\end{enumerate}

A \define{labelled 1-path}{}, $lP(V, E, L_V, L_E, \pi_0, \pi_1, l_V, l_E)$, is a labeled 
graph whose underlying graph is a 1-path.
\end{definition}

Given that $1$-paths are total orders (and hence are posets), we \emph{could} use the more
traditional, ``$\leq$'', notation for the edges in $1$-paths. However to ensure a more
obvious generalisation to $\lambda$-paths (for any ordinal, $\lambda$), we retain the edge
notation, ``$(x,y)$'', throughout.

\begin{definition}
For a path, $P(V, E, \pi_0, \pi_1)$, a vertex, $v \in V$, is \define{maximal}{} or
\define{final}{} (respectively \define{minimal}{} or \define{initial}{}) if for every
vertex, $x \in V$, there exists an edge $(x,v) \in E$ ($(v, x) \in E$).

A vertex is \define{internal}{} if it is neither initial or final. A vertex is on the  
\define{boundary}{} if it is either initial or final.

A \define{$1$-interval}{} is a $1$-path which has both maximal and minimal vertices. Given 
verticies, $x, y \in V$ which are minimal and maximal (respectively), we often denote the 
interval as, $[x,y]$.
\end{definition}

Note that by anti-symmetry, any maximal (minimal) elements of a path are unique. By total
order, if there exists two distinct vertices, $x,y \in V$, for which $(x,y) \in E$, then
the vertex $x$ can not be maximal, and conversely, the vertex $y$ can not be minimal.

\begin{definition}
A path, $P_0(V^{P_0}, E^{P_0}, \pi^{P_0}_0, \pi^{P_0}_1)$, is a \define{subpath}{} of a
path, $P_1(V^{P_1}, E^{P_1}, \pi^{P_1}_0, \pi^{P_1}_1)$, if there exists an embedding,
\map{\phi}{P_0}{P_1}.

Conversely, a path, $P_1(V^{P_1}, E^{P_1}, \pi^{P_1}_0, \pi^{P_1}_1)$, \define{extends}{}
a path, $P_0(V^{P_0}, E^{P_0}, \pi^{P_0}_0, \pi^{P_0}_1)$, if there exists an embedding,
\map{\phi}{P_0}{P_1}.
\end{definition}

Note that every edge in any path is trivially a subpath of the path. Conversely, a path
extends each of its edges. By abuse of notation, the edge, $(x,y)$, will often also denote
its corresponding sub\emph{path}, $P(\set{x,y}, \set{(x,y)}, \pi_0((x,y)):=x,
\pi_1((x,y)):=y)$.

\begin{definition}
Given paths, $P_0(V^{P_0}, E^{P_0}, \pi^{P_0}_0, \pi^{P_0}_1)$ and $P_1(V^{P_1}, E^{P_1},
\pi^{P_1}_0, \pi^{P_1}_1)$, an edge, $(x^{P_0}, y^{P_0}) \in E^{P_0}$ \define{covers}{} an
edge, $(x^{P_1}, y^{P_1}) \in E^{P_1}$, if there exists an subpath, $\hat{P_1}$ in $P_1$
which is both an interval, $[\hat{x^{P_1}}, \hat{y^{P_1}}]$, and an extension of
$(x^{P_1}, y^{P_1})$ for which there exists an embedding, \map{\phi}{(x^{P_0},
y^{P_0})}{\hat{P_1}} for which:
\begin{itemize}
\item $\phi(x^{P_0}) = \hat{x^{P_1}}$
\item $\phi(y^{P_0}) = \hat{y^{P_1}}$
\item $\phi( (x^{P_0}, y^{P_0}) ) = (\hat{x^{P_1}}, \hat{y^{P_1}})$
\end{itemize}
\end{definition}

\begin{definition}
A path, $P_1(V^{P_1}, E^{P_1}, \pi^{P_1}_0, \pi^{P_1}_1)$, is a \define{refinement}{} of a
path, $P_0(V^{P_0}, E^{P_0}, \pi^{P_0}_0, \pi^{P_0}_1)$, if every edge, $e^{P_1} \in 
E^{P_1}$ is covered by an edge $e^{P_0} \in E^{P_0}$.
\end{definition}

Having defined what a $1$-path is, we now need to explore the collection of $1$-paths.

\begin{theorem}\label{path:addFinal}
Given a path, $P(V, E, \pi_0, \pi_1)$, there exists a path, $P^+(V^+, E^+, \pi^+_0,
\pi^+_1)$,  unique up to graph isomorphism, for which
\begin{enumerate}
\item there exist a graph embedding, \map{\phi^+}{P}{P^+}
\item $V^+ = \phi^+(V) \union \set{V}$
\item for each $v \in V$ there exists an edge, $(\phi^+(v), \set{V}) \in E^+$.
\end{enumerate}
\end{theorem}

\begin{proof}
\TODO{PROVE THIS!}
\end{proof}

\begin{corollary}\label{path:addInitial}
Given a path, $P(V, E, \pi_0, \pi_1)$, there exists a path, $P^-(V^-, E^-, \pi^-_0,
\pi^-_1)$,  unique up to graph isomorphism, for which
\begin{enumerate}
\item there exist a graph embedding, \map{\phi^-}{P}{P^-}
\item $V^- = \phi^-(V) \union \set{V}$
\item for each $v \in V$ there exists an edge, $(\set{V}, \phi^-(v)) \in E^-$.
\end{enumerate}
\end{corollary}

\begin{proof}
Apply Theorem, \ref{path:addFinal}, to the dual path of $P$.
\end{proof}

\begin{corollary}
Given a path, $P(V, E, \pi_0, \pi_1)$, there exists a path, $P^{+-}(V^{+-}, E^{+-},
\pi^{+-}_0, \pi^{+-}_1)$,  unique up to graph isomorphism, for which
\begin{enumerate}
\item there exist a graph embedding, \map{\phi^{+-}}{P}{P^{+-}}
\item $V^{+-} = \phi^{+-}(V) \union \set{V} \union \set{V^+}$
\item for each $v \in V$ there exists an edge, $(\phi^{+-}(v), \set{V}) \in E^{+-}$,
\item for each $v \in V$ there exists an edge, $(\set{V^+}, \phi^{+-}(v)) \in E^{+-}$,
\item $(\set{V^+}, \set{V}) \in E^{+-}$.
\end{enumerate}
\end{corollary}

\begin{proof}
Apply Theorem, \ref{path:addFinal}, to obtain the path $P^+(V^+, E^+, \pi^+_0, \pi^+_1)$
which has a unique final vertex, and then Corollary, \ref{path:addInitial}, to obtain
$P^{+-}$.
\end{proof}

\begin{theorem}
Given a pair of paths, $P^-(V^-, E^-, \pi^-_0, \pi^-_0)$ and $P^+(V^+, E^+, \pi^+_0,
\pi^+_1)$, for which $P^-$ has a final vertex, $p^-$ and $P^+$ has an initial vertex,
$p^+$, there exists a path $P(V, E, \pi_0, \pi_1)$, unique up to graph isomorphism, for
which
\begin{enumerate}
\item there exists a graph embeddings, \map{\phi^-}{P^-}{P} and \map{\phi^+}{P^+}{P},
\item $\phi^-(p^-) = \phi^+(p^+)$,
\item for each $v^+ \in V^+$ and $v^+ \in V^+$ there exists an edge $(\phi^-(v^-), 
\phi^+(v^+)) \in E$.
\end{enumerate}
\end{theorem}

\begin{proof}
\TODO{PROVE THIS!}
\end{proof}

\TODO{Add associativity of path joining}

\begin{definition}
A labelled path, $lP(V, E, L_V, L_E, \pi_0, \pi_1, l_V, l_E)$, is \define{edge
consistent}{} if the set of edge labels, $L_E$, is equipped with a mapping,
\map{\compose}{L_E \times L_E}{L_E}, for which if $x, y, z \in V$ are three verticies for 
which $(x,y), (y,z) \in E$ (and hence $(x,z) \in E$), then $l_E((x,z)) = l_E((x,y)) \compose 
l_E((y,z))$.
\end{definition}

\begin{definition}
An graph embedding, \map{\phi}{P}{lG}, of a path, $P(V^P, E^P, \pi^P_0, \pi^P_1)$ into a 
labelled graph, $lG(V^{lG}, E,^{lG}, L_{V^{lG}}, L_{E^{lG}}, \pi^{lG}_0, \pi^{lG}_1, 
l_{V^{lG}}, l_{E^{lG}})$ induces a labelling on the path, $P$, via the pullback along 
$\phi$. Specifically, the \define{induced labelling}{} is the labelled path, $lP(V^P, E^P, 
L_{V^{lG}}, L_{E^{lG}}, \pi^P_0, \pi^P_1, l_{V^{lG}} \compose \phi, l_{E^{lG}} \compose 
\phi)$
\end{definition}

\begin{definition}
A labelled graph, $lG(V, E, L_V, L_E, \pi_0, \pi_1, l_V, l_E)$, has an \define{extensional
edge labelling}{} if two edges, $e_0, e_1 \in E$ are equal, $e_0 =_E e_1$ (as elements of
the set $E$) if and only if
\begin{enumerate}
\item $\pi_0(e_0) =_V \pi_0(e_1)$
\item $\pi_1(e_0) =_V \pi_1(e_1)$
\item $l_E(e_0) =_{L_E} l_E(e_1)$
\end{enumerate}
as elements of the respective sets.
\end{definition}

For at most countable graphs, defining ``linearity'' is relatively easily done by
``following'' single edges corresponding to the ``successor'' or ``next'' operation. For
trans-finite graphs we must add something more to be able to deal with ``limit points''.
Rather surprisingly, to deal with these ``limit points'' it is convenient to define
1-Categories, since a Category which is, sequentially complete as well as sequentially
cocomplete, is a graph which explicitly includes all of its ``paths''. Categorically, a
category is a presheaf on the category
$ \bullet \substack{\mathbf{\longrightarrow} \\[-0.7ex] \mathbf{\longleftarrow} 
\\[-0.7ex] \mathbf{\longrightarrow} } \bullet $
equipped with a consistent collection of 2-simplicies which are declared
``thin''\footnote{This way of looking at 1-categories is inspired by the work of Dominic
Verity, see, for example, \cite{verity2005complicialSets},
\cite{verity2006complicialSimplicialHomotopy} and
\cite{verity2006simplicialComplicialCategories}.}.

\begin{definition}
A \define{1-category}{}, $C(V, E, L_V, L_E, \pi_0, \pi_1, l_V, l_E, i)$, is a labelled graph 
with an extensional edge labelling, \map{l_E}{E}{L_E}, 
\end{definition}


\begin{definition}
A \define{1-category}{}, $C(E, V, \pi_0, \pi_1, i)$, is a graph with the addition of an 
\define{identity map}{} \map{i}{V}{E} for which $\pi_0 \compose i$ and $\pi_1 
\compose i$ are both the identity of the set $V$, and for which 
\begin{itemize}
\item If $f, g \in E$ and $\pi_1(f) = \pi_0(g)$ then there exists $f \compose g \in E$ for 
which $ \pi_0(f) = \pi_0(f \compose g)$ and $ \pi_1(f \compose g) = \pi_1(g)$
\item If $f \in E$ then $ f \compose i(\pi_0(f)) = f$ and $i(\pi_1(f)) \compose f = f$
\item If \map{f}{w}{x}, \map{g}{x}{y} and \map{h}{y}{z} then the graph, $hgf(\set{f,
g, h}, \set{w, x, y, z}, \pi_0, \pi_1)$ can be extended to a 1-path embedded in $C$ in only 
one way
\end{itemize}
\end{definition}
