% LaTeX source for the trans-finite Turing document
%

%\documentclass[a4paper,openany]{amsbook}
\documentclass[a4paper]{amsart}
\usepackage{disitt}
\usepackage{disitt-symbols}
\usepackage{mdframed}
\newmdenv[linecolor=white,backgroundcolor=gray!10]{infobox}
\newenvironment{myQuote}{\begin{quotation}}{\end{quotation}}
\surroundwithmdframed[linecolor=white,backgroundcolor=gray!10]{myQuote}

\begin{document}
%\frontmatter
\sloppy
% see: http://tex.stackexchange.com/a/50850
%\hfuzz=20pt
%\vfuzz=20pt
%\hbadness=10000
%\vbadness=\maxdimen

\title[Computation: Classical version]{Trans-finite Computation: The classical version}
\input{frontMatter}
\subjclass[2010]{Primary unknown; Secondary unknown} %
\keywords{Keyword one, keyword two etc.}%

\begin{abstract}
We investigate what it means for a finite being to verify trans-finite computational
algorithms.
\end{abstract} 
\maketitle 
\tableofcontents 
%\mainmatter


\section{Introduction}

What is a computation? Turing, \cite{turing1936computableNumbersMachines}, is cited by
G\"odel, \cite{godel}, as providing the first convincing \emph{mechanistic} mathematical
definition of a computational process as performed by a human. Turing's student, Gandy,
\cite{gandy1980churchsThesisMechanisms}, provided a similar \emph{mechanistic}
mathematical definition of a computational process as performed by a \emph{finite
machine}. Both Turing and Gandy focused upon providing arguments justifying why particular
mathematical structures provided reasonable definitions of computation. In order to
understand potentially trans-finite computation, we will take an essentially dual
approach. We will look at mathematical structures which represent \emph{traces} of a
(trans-finite) computation, and then work back to mathematical structures which could have
produced these traces. Rather surprisingly, to capture \emph{trans-finite} computational
traces, we are forced investigate the \emph{categorical} structures of labelled graphs.

What is a computation? Any \emph{description} of a computational process made by a finite
being, will inevitably consist of a collection of ``marks in the sand'', ink-marks on
paper, or bit-patterns in the memory of a computer. Theoretical Computer Science has three
ways of verifying properties of the computational processes represented by these
collections of symbols. 
%
\begin{itemize}
%
\item \define{Operational Semantics}{},
\cite{plotkin1981structuralOperationalSemantics}, provides a description of a transition 
system which models a program.
%
\item \define{Denotational
Semantics}{}, \cite{scott1970theoryComputation, stoy1981scottStracheyTheory}, provides
mathematical structures which represent the underlying \emph{state} of a compuation.
%
\item \define{Axiomatic Semantics}{}, \cite{floyd1967programSemantics, 
hoare1969axiomaticSemantics} provides an axiomatic system of natural deduction rules with 
which to \emph{prove} that a program meets a specific \emph{logical} specification.
%
\end{itemize}
%
Winskel, \cite{winskel1993formalSemanticsProgrammingLanguages}, provides a good
introduction to these three types of computational semantics\footnote{Not surprisingly the
ideas contained in this article were originally inspired by Winskel's study of
computational traces using event structures and bisimulation, see, for example,
\cite{winskel2007eventStructures}, and \cite{joyalNielsenWinskel1996bisimulation}}. Gries,
\cite{gries1981scienceProgramming}, provides a particularly detailed account of building
programs via their axiomatic semantical proofs of correctness.

What is a computation? A \define{computation}{} is the transformation of one mathematical
structure representing an un-transformed \define{state}{} into a correspondingly modified
mathematical structure representing the transformed state. Essentially the genius of
Turing and Gandy was to define those transformations which a finite human, or machine
could perform. What happens if we place no such restrictions? If there are no
restrictions, how do we know what a computation computes?

\TODO{merge these different introductions}

This paper is a prequel to our cycle of papers aiming to provide a rigorous mathematical
theory of what limited, ``Modi-potent'', beings can understand of Reality. One of the
central tenets of this ``Theory of Reality'' is that the correct foundations of
Mathematics is not first order logic but rather computation. However to recover all of
classical mathematics inside these computational foundations, we need a theory of
computation which is more powerful than classical (non oracle) Turing machines allow. This
paper provides definitions of these more powerful, trans-finite Turing machines in a
classical Set Theory setting. Subsequent papers will redo these definitions in the context
of computational foundations.

There have been numerous previous attempts to define ``more powerful'' Turing machines,
some of which, for example, oracle or ``o-machines'', date back to Turing's original work.
More recently the area of ``hypercomputation'' has attracted some interest, some of which
is rather controversial. Ord, \cite{ord2006hyperComputation}, provides a good review of
much of this work\footnote{For a high-level overview with some additional citations, see
the Wikipedia articles, \cite{wikipedia2015hyperComputation} and
\cite{wikipedia2015superRecursiveAlgorithm}, as well as the nLab article,
\cite{nLab2015hyperComputation}}. Some authors claim that there (may) exist physical
implementations of hypercomputational devices, \emph{we make \emph{no} such claims}.
Davis, \cite{davis2004hyperComputation}, provides a summary of some of the arguments
against these claims. Instead, our interest comes from our desire to capture
non-constructive classical mathematics in a computational setting.

\TODO{\cite{hamkinsLewis1998inifiniteTimeTuringMachines}}

Our aim in this paper is to explicitly explore a more uniform definition of Turing machine
which will differ simply in the power of the maximal ordinal used by the Turing machine.
We do this using ``classical'' socially checked proofs using ``classical'' set theory. You
are welcome to choose your favourite axiomization using first order predicate logic. In
subsequent papers we will found mathematics using a form of trans-finite computation which
is equivalent to our current definition of trans-finite Turing Machine. 

Classically, the interest and hence the definitions of both computation and set theory,
have been focused upon computations which \emph{halt} and sets which are
\emph{well-founded}. Our computational foundations will admit that there are some
interesting computations and structures which do not halt or, alternatively, are
non-well-founded. In categorical terms the halting computations and well-founded sets are
\emph{algebraic}, while the non-halting processes and non-well-founded structures are
\emph{co-algebraic}. While we will not make any use of co-algebraic structures in this
paper, our subsequent papers will make essential use of these co-algebraic structures.

Of particular importance for either the classical or computational definitions of
trans-finite computation, is the power of the Axiom of Choice(s) used. For our purposes
there are three versions of the Axiom of Choice, in increasing power:
%
\begin{enumerate}
\item Prime Ideal Theorem (PIT)
\item Axiom of Choice (AoC)
\item Global Axiom of Choice (GAoC)
\end{enumerate}

The Prime Ideal Theorem (PIT) is used to define the ``result'' of computation at
limit ordinals, which corresponds to creation of limit points or alternatively
the closure, via ``ideal points'' of any computation.

The Axiom of Choice (AoC) itself, is used to define sequential computation, or
equivalently, to ensure the existence of lists of any required ordinal size.

While we will not use the Global Axiom of Choice (GAoC) in any essential way in
our work, we will see that the GAoC is required to ensure the existence of
co-algebraic lists, when we work, in subsequent papers, with Co-Algebraic
structures.

\section{Time: Graphs, paths and refinements}

In anticipation of our future work, we take a slightly non-standard route towards defining
(trans-finite) computation. We leave it to the reader to understand both why we take
this route, and that we have, in fact, provided the same structures.

The classical theory of computation, with its (almost exclusive) focus on
\emph{terminating} computational processes, is largely focused on \emph{time} to the
exclusion of \emph{space}. For \emph{terminating} processes, any parallel (spacial)
process can be serialized. This serialisability \emph{is} equivalent to set theory's
Well-ordering Theorem and hence the Axiom of Choice. As we will see below,
\emph{non-terminating} processes need not be globally serializable. This is, again,
equivalent to the lack of global time in relativistic space-time.

Since we are not omni-potent beings, while we can prove theorems about non-terminating 
processes, we can not directly experience them. We can only access non-terminating processes 
via (terminating) ordinal approximations to the full non-terminating process. This means 
that, in a sense, ``all'' computation can \emph{effectively} be reduced to serializable 
terminating processes. 

In this section we explore the \emph{time} focused foundations of computations. In the next 
section we explore the addition of \emph{space-time} (and hence parallelism).

One of the central themes of this cycle of papers is the importance of graphs, paths and
(co-)algebraic refinements. We start by defining (trans-finite) graphs.

There are essentially two distinct ways to define a mathematical structure: logically, by
explicitly assuming the existence of a set with specific (logically expressed) properties,
or computationally, in categorical language, as the fixed point of an endofunctor. From a
non-categorical inductive point of view, an early expression of this dichotomy can be
found in \cite[page 1]{moschovakis1974induction}. In this paper we follow the logical
approach. In subsequent papers in this cycle, we will use the computational approach.

There are relatively few discussions of non-finite graphs from which we mention Reinhard
Diestel's work, \cite{diestel2006graphTheory, diestel1990infiniteGraphTheory}.
Unfortunately, for trans-finite graphs, we must go, more or less, ``on beyond
zebra''\footnote{Growing up, one of my favourite books was Dr Seuss' ``On beyond Zebra'',
\cite{seuss1955onBeyondZebra}. \emph{Be warned}: One of my greatest faults is that I seem
congenitally incapable of staying within ``obvious'' bounds...}\footnote{Another of my
greatest faults is that my mind is a web of snippets of ideas. Linearizing this web into
single narrative, is \emph{very} hard for me to do, and may be hard to fully follow until 
you have heard ``the whole shaggy-dog story''.}.

Our definition of graph is inspired by the definition of metagraph in
\cite{macLane1971categoriesWorkingMathematician} section I.1. Note that our definition is
a generalisation of most typical definitions, for example \cite[section
1.1]{diestel2006graphTheory}. In particular, all of our graphs are \emph{explicilty}
\emph{directed}. Categorically, a graph is simply a presheaf on the category
$ \cdot \substack{\mathbf{\longrightarrow} \\[-0.7ex] \mathbf{\longrightarrow}} \cdot $

\begin{definition}
A \define{graph}{}, $G(V, E, \pi_0, \pi_1)$, is a set, $V$, of \define{verticies}{}, and a
set, $E$, of \define{edges}{}, together with two maps, \map{\pi_0}{E}{V} and
\map{\pi_1}{E}{V}, called the source and target maps.

A \define{labeled graph}{}, $lG(V, E, L_V, L_E, \pi_0, \pi_1, l_V, l_E)$, is a graph $G(V,
E, \pi_0, \pi_1)$ together with a set, $L_V$, of \define{vertex labels}{}, and a set,
$L_E$, of \define{edge labels}{}, and maps, \map{l_V}{V}{L_V}, and \map{l_E}{E}{L_E}.
Often the set of vertex labels, $L_V$, and vertex label map, $l_V$, are omitted (or
alternatively the set of vertex labels is assumed to be the trivial one point set).
 
Given a labeled graph, $lG(V, E, L_V, L_E, \pi_0, \pi_1, l_v, lE)$ its \define{underlying
graph}{}, is the graph, $G(V, E, \pi_0, \pi_1)$.
\end{definition}

\begin{definition}
A \define{graph morphism}{}, \map{\phi}{G_0}{G_1}, is a map between the graphs,
$G_0(V^{G_0}, E^{G_0}, \pi^{G_0}_0, \pi^{G_0}_1)$ and $G_1(V^{G_1}, E^{G_1}, \pi^{G_1}_0,
\pi^{G_1}_0)$ for which
\begin{enumerate}
\item $\phi(V^{G_0}) \subset V^{G_1}$
\item $\phi(E^{G_0}) \subset E^{G_1}$
\item $\phi \compose \pi^{G_0}_i = \pi^{G_1}_i \compose \phi$ for $i \in {0, 1}$
\end{enumerate}

A \define{labeled graph morphism}{}, \map{\phi}{lG_0}{lG_1}, between a pair of labeled
graphs, $lG_0(V^{lG_0}, E^{lG_0}, L^{lG_0}_V, L^{lG_0}_E, \pi^{lG_0}_0, \pi^{lG_0}_1,
l^{lG_0}_V, l^{lG_0}_V)$ and $lG_1(V^{lG_1}, E^{lG_1}, L^{lG_1}_V, L^{lG_1}_E,
\pi^{lG_1}_0, \pi^{lG_1}_1, l^{lG_1}_V, l^{lG_1}_V)$, is a graph morphism between the
underlying graphs, $G_0$ and $G_1$ for which
\begin{enumerate}
\item $\phi(L^{lG_0}_V) \subset L^{lG_1}_V$
\item $\phi(L^{lG_0}_E) \subset L^{lG_1}_E$
\item $\phi \compose l^{lG_0}_V = l^{lG_1}_V \compose \phi$
\item $\phi \compose l^{lG_0}_E = l^{lG_1}_E \compose \phi$
\end{enumerate}
\end{definition}

\begin{definition}
A \define{graph embedding}{} is a graph morphism, \map{\phi}{G_0}{G_1}, respectively, a
labelled graph morphism, \map{\phi}{lG_0}{lG_1}, which is, categorically, a monomorphism.
\end{definition}

\begin{lemma}
Given a graph, $G(V, E, \pi_0, \pi_1)$, there exists a \define{dual graph}{},
$\tilde{G}(\tilde{V}, \tilde{E}, \tilde{\pi}_0, \tilde{\pi}_1)$, unique up to graph 
isomorphism, for which
\begin{enumerate}
\item $\tilde{V} = V$,
\item $\tilde{E} = \set{(v, u) \suchThat (u, v) \in E}$
\end{enumerate}
\end{lemma}

\begin{proof}
\TODO{PROVE THIS!}
\end{proof}

Our next goal is to provide a definition of $1$-dimensional (time) ``$1$-paths''.
Intuitively a (directed) ``$1$-path'' \emph{in} a graph is an embedding of a ``totally
ordered'' directed graph.

\begin{definition}
A \define{(directed) 1-path}{}, $P(V, E, \pi_0, \pi_1)$, is a graph for which:

\begin{enumerate}
\item \define{Anti-symmetry:}{} for any pair of vertices, $x, y \in V$, if $(x,y) \in E$ and
$(y,x) \in E$ then $x = y$.
\item \define{Transitivity:}{} for any triple of verticies, $x, y, z \in V$, 
if $(x,y) \in E$ and $(y,z) \in E$ then $(x,z) \in E$.
\item \define{Total order:}{} for any pair of vertices, $x, y \in V$, if $x \neq y$ then
either $(x,y) \in E$ or $(y,x) \in E$ but not both.
\end{enumerate}
\end{definition}

Given that $1$-paths are total orders (and hence are posets), we \emph{could} use the more
traditional, ``$\leq$'', notation for the edges in $1$-paths. However to ensure a more
obvious generalisation to $\lambda$-paths (for any ordinal, $\lambda$), we retain the edge
notation, ``$(x,y)$'', throughout.

\begin{definition}
For a path, $P(V, E, \pi_0, \pi_1)$, a vertex, $v \in V$, is \define{maximal}{} or
\define{final}{} (respectively \define{minimal}{} or \define{initial}{}) if for every
vertex, $x \in V$, there exists an edge $(x,v) \in E$ ($(v, x) \in E$).

A vertex is \define{internal}{} if it is neither initial or final. A vertex is on the  
\define{boundary}{} if it is either initial or final.

A \define{$1$-interval}{} is a $1$-path which has both maximal and minimal vertices. Given 
verticies, $x, y \in V$ which are minimal and maximal (respectively), we often denote the 
interval as, $[x,y]$.
\end{definition}

Note that by anti-symmetry, any maximal (minimal) elements of a path are unique. By total
order, if there exists two distinct vertices, $x,y \in V$, for which $(x,y) \in E$, then
the vertex $x$ can not be maximal, and conversely, the vertex $y$ can not be minimal.

\begin{definition}
A path, $P_0(V^{P_0}, E^{P_0}, \pi^{P_0}_0, \pi^{P_0}_1)$, is a \define{subpath}{} of a
path, $P_1(V^{P_1}, E^{P_1}, \pi^{P_1}_0, \pi^{P_1}_1)$, if there exists an embedding,
\map{\phi}{P_0}{P_1}.

Conversely, a path, $P_1(V^{P_1}, E^{P_1}, \pi^{P_1}_0, \pi^{P_1}_1)$, \define{extends}{}
a path, $P_0(V^{P_0}, E^{P_0}, \pi^{P_0}_0, \pi^{P_0}_1)$, if there exists an embedding,
\map{\phi}{P_0}{P_1}.
\end{definition}

Note that every edge in any path is trivially a subpath of the path. Conversely, a path
extends each of its edges. By abuse of notation, the edge, $(x,y)$, will often also denote
its corresponding sub\emph{path}, $P(\set{x,y}, \set{(x,y)}, \pi_0((x,y)):=x,
\pi_1((x,y)):=y)$.

\begin{definition}
Given paths, $P_0(V^{P_0}, E^{P_0}, \pi^{P_0}_0, \pi^{P_0}_1)$ and $P_1(V^{P_1}, E^{P_1},
\pi^{P_1}_0, \pi^{P_1}_1)$, an edge, $(x^{P_0}, y^{P_0}) \in E^{P_0}$ \define{covers}{} an
edge, $(x^{P_1}, y^{P_1}) \in E^{P_1}$, if there exists an subpath, $\hat{P_1}$ in $P_1$
which is both an interval, $[\hat{x^{P_1}}, \hat{y^{P_1}}]$, and an extension of
$(x^{P_1}, y^{P_1})$ for which there exists an embedding, \map{\phi}{(x^{P_0},
y^{P_0})}{\hat{P_1}} for which:
\begin{itemize}
\item $\phi(x^{P_0}) = \hat{x^{P_1}}$
\item $\phi(y^{P_0}) = \hat{y^{P_1}}$
\item $\phi( (x^{P_0}, y^{P_0}) ) = (\hat{x^{P_1}}, \hat{y^{P_1}})$
\end{itemize}
\end{definition}

\begin{definition}
A path, $P_1(V^{P_1}, E^{P_1}, \pi^{P_1}_0, \pi^{P_1}_1)$, is a \define{refinement}{} of a
path, $P_0(V^{P_0}, E^{P_0}, \pi^{P_0}_0, \pi^{P_0}_1)$, if every edge, $e^{P_1} \in 
E^{P_1}$ is covered by an edge $e^{P_0} \in E^{P_0}$.
\end{definition}

Having defined what a $1$-path is, we now need to explore the collection of $1$-paths.

\begin{theorem}\label{path:addFinal}
Given a path, $P(V, E, \pi_0, \pi_1)$, there exists a path, $P^+(V^+, E^+, \pi^+_0,
\pi^+_1)$,  unique up to graph isomorphism, for which
\begin{enumerate}
\item there exist a graph embedding, \map{\phi^+}{P}{P^+}
\item $V^+ = \phi^+(V) \union \set{V}$
\item for each $v \in V$ there exists an edge, $(\phi^+(v), \set{V}) \in E^+$.
\end{enumerate}
\end{theorem}

\begin{proof}
\TODO{PROVE THIS!}
\end{proof}

\begin{corollary}\label{path:addInitial}
Given a path, $P(V, E, \pi_0, \pi_1)$, there exists a path, $P^-(V^-, E^-, \pi^-_0,
\pi^-_1)$,  unique up to graph isomorphism, for which
\begin{enumerate}
\item there exist a graph embedding, \map{\phi^-}{P}{P^-}
\item $V^- = \phi^-(V) \union \set{V}$
\item for each $v \in V$ there exists an edge, $(\set{V}, \phi^-(v)) \in E^-$.
\end{enumerate}
\end{corollary}

\begin{proof}
Apply Theorem, \ref{path:addFinal}, to the dual path of $P$.
\end{proof}

\begin{corollary}
Given a path, $P(V, E, \pi_0, \pi_1)$, there exists a path, $P^{+-}(V^{+-}, E^{+-},
\pi^{+-}_0, \pi^{+-}_1)$,  unique up to graph isomorphism, for which
\begin{enumerate}
\item there exist a graph embedding, \map{\phi^{+-}}{P}{P^{+-}}
\item $V^{+-} = \phi^{+-}(V) \union \set{V} \union \set{V^+}$
\item for each $v \in V$ there exists an edge, $(\phi^{+-}(v), \set{V}) \in E^{+-}$,
\item for each $v \in V$ there exists an edge, $(\set{V^+}, \phi^{+-}(v)) \in E^{+-}$,
\item $(\set{V^+}, \set{V}) \in E^{+-}$.
\end{enumerate}
\end{corollary}

\begin{proof}
Apply Theorem, \ref{path:addFinal}, to obtain the path $P^+(V^+, E^+, \pi^+_0, \pi^+_1)$
which has a unique final vertex, and then Corollary, \ref{path:addInitial}, to obtain
$P^{+-}$.
\end{proof}

\begin{theorem}
Given a pair of paths, $P^-(V^-, E^-, \pi^-_0, \pi^-_0)$ and $P^+(V^+, E^+, \pi^+_0,
\pi^+_1)$, for which $P^-$ has a final vertex, $p^-$ and $P^+$ has an initial vertex,
$p^+$, there exists a path $P(V, E, \pi_0, \pi_1)$, unique up to graph isomorphism, for
which
\begin{enumerate}
\item there exists a graph embeddings, \map{\phi^-}{P^-}{P} and \map{\phi^+}{P^+}{P},
\item $\phi^-(p^-) = \phi^+(p^+)$,
\item for each $v^+ \in V^+$ and $v^+ \in V^+$ there exists an edge $(\phi^-(v^-), 
\phi^+(v^+)) \in E$.
\end{enumerate}
\end{theorem}

\begin{proof}
\TODO{PROVE THIS!}
\end{proof}

For at most countable graphs, defining ``linearity'' is relatively easily done by
``following'' single edges corresponding to the ``successor'' or ``next'' operation. For
trans-finite graphs we must add something more to be able to deal with ``limit points''.
Rather surprisingly, to deal with these ``limit points'' it is convenient to define
1-Categories, since a Category which is, sequentially complete as well as sequentially
cocomplete, is a graph which explicitly includes all of its ``paths''. Categorically, a
category is a presheaf on the category
$ \bullet \substack{\mathbf{\longrightarrow} \\[-0.7ex] \mathbf{\longleftarrow} 
\\[-0.7ex] \mathbf{\longrightarrow} } \bullet $
equipped with a consistent collection of 2-simplicies which are declared
``thin''\footnote{This way of looking at 1-categories is inspired by the work of Dominic
Verity, see, for example, \cite{verity2005complicialSets},
\cite{verity2006complicialSimplicialHomotopy} and
\cite{verity2006simplicialComplicialCategories}.}.

\begin{definition}

A \define{1-category}{}, $C(E, V, \pi_0, \pi_1, i)$, is a graph with the addition of an 
\define{identity map}{} \map{i}{V}{E} for which $\pi_0 \compose i$ and $\pi_1 
\compose i$ are both the identity of the set $V$, and for which 

\begin{itemize}

\item If $f, g \in E$ and $\pi_1(f) = \pi_0(g)$ then there exists $f \compose g \in E$ for 
which $ \pi_0(f) = \pi_0(f \compose g)$ and $ \pi_1(f \compose g) = \pi_1(g)$

\item If $f \in E$ then $ f \compose i(\pi_0(f)) = f$ and $i(\pi_1(f)) \compose f = f$

\item If \map{f}{w}{x}, \map{g}{x}{y} and \map{h}{y}{z} then the graph, $hgf(\set{f,
g, h}, \set{w, x, y, z}, \pi_0, \pi_1)$ can be extended to a 1-path embedded in $C$ in only 
one way

\end{itemize}

\end{definition}

\section{Space-time: directed simplicies, refinements and space-time paths}

In the previous section we explored the structures required to provide a \emph{time} focused 
foundations of computation. In this section we expand these time focused structures into 
their corresponding fully space-time structures.

\section{State machines}

\begin{definition}

A state machine... A finite state machine is a state machine ...

\end{definition}

\subsection{$\aleph_0$-non-deterministic Turing Machine}

\subsection{o-machines}

\section{Trans-finite defintions}

\section{Relationship with o-machines}

\begin{theorem}

Given a set, $\gamma$, there exists a trans-finite Turing machine, $T$, corresponding to
the $\gamma$-o-machine, $O$.

\end{theorem}


\begin{theorem}

Given a trans-finite Turing machine, $T$, there exists a set, $\gamma$, and a
$\gamma$-o-machine, $O$, that simulates $T$.

\end{theorem}

\begin{proof}
something proving the previous theorem.
\end{proof}

\section{Trans-finite computability theory}


\bibliographystyle{amsalpha}
\bibliography{transFiniteComputation}

\end{document}

