% A ConTeXt document [master document: hilbertsProgram.tex]

\section[title=Some Philosophy]

In Western philosophy, since at least the time of the ancient Greeks, 
there have been a wide range of Philosophical theories of the 
\quote{Reality} of \quote{Reality}. Our objective in building a rigorous 
Mathematical theory of Reality is not to prove any of these Philosophical 
theories (in)correct. Instead our objective, and really the only one 
available \emph{Mathematically}, is to explore what a finite computational 
device, \emph{\quote{a wee beastie}}, can learn about Reality. 

\subsection[title=Hume's problem]

Ignoring Hume's \quote{Problem of Induction} for the moment, as a 
Scientist and Engineer, like any young child, I live in the belief that I 
can both learn about and, more importantly, \emph{interact} with Reality. 
To bastardize Descartes, \emph{from moment to moment, I can see that I 
have made marks in the sand, therefore I am}. 

It is naive to assert that finite beings, such as ourselves, can not learn 
to predict at least some of the future. Russell's farmyard birds, given 
their limited cognitive abilities, \emph{are rational} to expect to be 
feed daily\footnote{See chapter VI, \quote{On Induction}, in Russell's 
\quote{The Problems of Philosophy}, \cite{russell1912problemsOfPhilosophy} 
near page 98}. However, for any \emph{finite} being, there will always be 
events, some highly critical events, which are outside of that being's 
ability to know about and hence predict. This is emphatically the 
\emph{\quote{wee}} in my understanding of a \emph{\quote{wee beastie}}. 
For any finte being, there is always a more capable being, we might just 
not yet found these more capable beings. Understanding the limits of being 
finite, is the true import of Hume's Problem. 

It is equally naive to assert that a finite being can not interact with 
their environment. I \emph{can} communicate with you over both distance 
and time. We \emph{can} build (finite) computational devices. I am writing 
this document using one such device, you are no doubt reading this 
document using at least one other. So it is at least potentially 
reasonable to expect that \emph{finitist} Mathematics could be founded 
computationally. 

\TODO{What does it mean to \emph{exist} mathematically?}

\subsection[title=What is a thing?]

\blank[big]\startBlockQuote

\noindent From the range of the basic questions of metaphysics we shall here ask 
this one question: “What is a thing?” The question is quite old. What 
remains ever new about it is merely that it must be asked again and 
again.\footnote{Martin Heidegger, page 1, first paragraph, in 
\cite{heidegger1967whatIsAThing}, as quoted by 
\cite{doeringIsham2008thingTheoryFoundationsPhysics}.} 

\stopBlockQuote\blank[big]

A Zen Master would respond that there is \emph{no-thing}, there is only 
\emph{is-isness}, \emph{existence in its entirety}\footnote{Indeed a Zen 
Master would refuse to use mere words. To slightly mix philosophies, 
\quote{The Tao which can be named is not the Tao}, or again in Jewish 
tradition, God is not to be directly \emph{named}. Words differentiate 
\quote{things}, and Zen's \quote{existence-in-entirety}, the Tao and God 
are beyond all human limits to differentiate, identify or understand. We, 
as limited beings, can only \emph{experience}. This is similar to Cantor's 
expressed understanding of his Absolute Infinite Magnitudes, again, see 
Cantor's letter to Dedekind dated 1899, 
\cite{vanHeijenoort1967fregeToGodel}.}. That quarrelsome \quote{thing}, 
\emph{\quote{I}}, is only an illusion. The hardest thing any \quote{one} 
can do is to ignore the \emph{\quote{I}} in order to \quote{see} the 
\emph{\quote{is}}. The dissolution of this \emph{\quote{I}} is 
un-important in the context of the \emph{\quote{is}}. In our work, this 
point of view will be indispensable. 

However, for most of \quote{us}, such a view point is very hard to hold. 
We all play a \quote{me}-\quote{environment} game with existence. This 
view point is equally important for our work.

A \quote{quark} plays the absolute simplest of games, a 
\quote{quark}-\quote{everything-else} game. A \quote{quark}, re-acts to 
its environment. Any model of a \quote{quark} is a (fairly) simple 
S-Matrix. 

A frog's game is only slightly less simple, there is the frog, there are 
\quote{things} that are small enough to be potential food, there are 
\quote{things} that are so large they might be predators, and finally 
there are \quote{things} which might be potential mates. A frog's 
\quote{environment} has some substructure, \quote{prey}, 
\quote{predators}, and \quote{mates}. We assume that a frog's brain 
models, to a sufficiently complex level of detail, these three 
\quote{things}, however, by and large, frogs do not need to expend much 
more energy on playing any more complex games so they don't. By modelling 
only the most important categories, frogs can save energy by not building 
and maintaining complex and energetically expensive nervous 
systems\footnote{See, for example, Ewert's \emph{Motion Perception Shapes 
the Visual World of Amphibians}, 
\cite{ewert2004motionPerceptionAmphibians}. While this reference focuses 
primarily on the visual system of amphibians, it indicates an overall lack 
of need for complex models in an amphibian's nervous system. It is 
estimated, \cite{raichleGusnard2002brainEnergyBudget}, that the adult 
human brain consumes around 20\% of all calories consumed each day, yet 
the human brain only represents around 2\% of our body weight. Nervous 
systems \emph{are} relatively expensive to keep running. For most of a 
frog's needs, this additional complexity is not needed, this is a frog's 
evolutionary niche.} 

A human's game is \emph{much} more complex. We regularly, split 
\quote{our} environment into many many \quote{things}. \quote{Objects} for 
which we build wide classes of models of their behaviour and even their 
potential internal, \quote{intentional}, state of \quote{mind}. Chairs and 
mugs have different uses. Metals and glasses, have different abilities to 
be re-fashioned into useful tools. Animals have widely different 
behaviours providing useful companions or dangerous enemies. Even more 
complex, though, are our \quote{models} of other humans. Each person in 
our environment, has widely differing objectives of their \quote{own}. All 
of which we must, and, by and large, do, keep track of. For each of these 
objects we take an \emph{intentional stance}\footnote{See Daniel Dennett's 
\quote{The Intentional Stance}, \cite{dennett1987a}.}, they each have 
various, though widely, differing abilities to re-act, or intend with 
\quote{me}. These different intentional abilities are reflected in the 
overall complexity in the various models we build to represent any 
particular \quote{thing}. 

So how do we, most efficiently, build these models? Naively, we all 
\quote{know} what an object \quote{is} when we \quote{see} it. Physics 
suggests that all material things are made up of sub-atomic particles 
which are in turn made up of \quote{quarks}. It is \quote{obvious} at one 
level that \quote{I} and \quote{you} are \quote{different} things. 
However, when I shake your hand, where do \quote{my} \quote{quarks} end 
and \quote{your} \quote{quarks} begin? At the \quote{most basic} level, we 
can not separate one \quote{thing} from another \quote{thing}. A Zen 
master's view is actually deeply entwined with any complete mathematical 
model of \quote{things}. How do we reconcile these multiple levels of 
\quote{being} into one comprehensive and complete model of 
\quote{Reality}? 

Andreas D\"oring and Chris Isham in their paper, \emph{What is a thing?}, 
\cite{doeringIsham2008thingTheoryFoundationsPhysics}, suggest that Physics 
can best be captured using the variable sets point of view provided by 
Categorical Topos\footnote{See Lawvere's concept of variable sets, 
\cite{lawvere1975continuouslyVariableSets} or 
\cite{lawvereRosebrugh2003setsForMathematics}.}. For us, a Topos over a 
collection of descriptive \quote{levels}, provides the natural tool with 
which to capture the coherent variation of \quote{thing-ness} as our level 
of description varies. This suggests that the use of the structuralist 
Categorical point of view, in general, and the associated variable 
descriptive Topos point of view, in particular, will be very important to 
our work. 

\subsection[title=A Neuron's eye view]

So lets reflect for a moment on a neuron's point of view. When a neuron 
fires, it \quote{means} something, but what does it mean? Deeply embedded 
inside a complex collection of other neurons, each collecting the spike 
trains from countless other neurons. In some sense each neuron integrates 
the \quote{information} conveyed by each of these spike trains from 
up-stream neurons. What sort of information should these spike trains be 
communicating? At the very least they should be communicating some 
important value. This is what artificial neural networks model. 

\TODO{Bayesian brain, \cite{doyaIshiiPougetRao2007bayesianBrain}, Spikes, 
\cite{riekeWarlandDeRuyterVanSteveninck1999spikesNeuralCode} Pouget, 
\cite{beckPouget2007inferencesImplementationMarkov} and 
\cite{knillPouget2004bayesianUncertaintyComputation}, markov models, 
Shalizi spatial models, \cite{shalizi2001thesis}, section 10.2.1 \emph{Why 
Global States Are not Enough}, required. Markov has restrictive scope of 
description... but by recoding and using levels of description, we can 
recover the effect of history on the present and future.} 

\TODO{talk about Wally's oversight... \cite{walley1991impreciseProb}. 
There is a deep distinction between the ideal asymptotic reals and the 
finite subsequences of processes} 

\TODO{rework the following paragraphs}

So this paper generalizes the collected work of Spitters, Coquand, (see, 
for example, \cite{coquandSpitters2009integralsAndValuations})\footnote{It 
is also important to see the related work of Heunen, Landsman and 
Spitters, \cite{heunenLandsmanEtAl2009toposForAlgebraicQuantumTheory}}, 
together with Walley's work on imprecise probabilities, to produce a 
computable measure theory which a beastie could use. \TODO{Add the work of 
\cite{jackson2006phdThMeasureThSheaves} and its generalization in Isham 
\cite{doeringIsham2008thingTheoryFoundationsPhysics}, Section 8.2} 

As has become traditional in Imprecise Probability theory, in his book, 
\emph{Statistical Reasoning with Imprecise Probabilities}, 
\cite{walley1991impreciseProb}, Walley makes the upper and lower 
previsions (expectations) the primary objects of study with upper and 
lower probabilities as derived concepts. Since we will be concerned with 
probability based Markov structures we will reverse this orientation. One 
of the reasons Walley choose to work with previsions (expectations) 
instead of probabilities is because of his belief that Lower Probablities 
did not determine Lower Previsions (see section 2.7.3 page 82, 
\cite{walley1991impreciseProb}). In fact we will show below that with the 
correct definition of upper and lower measures and upper and lower 
integrals, lower probabilities do determine lower previsions. This will be 
the substance of the (Imprecise) Dedekind-Riesz Representation Theorms 
proven below. 

\TODO{paragraphs above}

\TODO{So what is a thing? A thing is an imprecise markov model at various 
levels of description. One of our ultimate goals is to provide a rigorous 
theory of computational imprecise measure theory sufficient to be able to 
provide the theory required to build computational \quote{wee beasties}.} 

\subsection[title=Mappae Mundi]

From our \quote{modern} point of view, Mappae Mundi are, at best, highly 
distorted (navigational) \quote{maps}. From the original 
\quote{cartographer's} point of view, each Mappa Mundi provides a useful 
representation of an important understanding about the 
\emph{cartographer's} \quote{world}. However, a modern notion of 
\quote{navigation} was highly \emph{un}-likely to be the primary purpose 
of any given historical Mappa Mundi. 

Research Mathematicians who habitually cross (standard) mathematical 
disciplines, of necessity, become collectors of mathematical mappa mundi. 
Each (matheamtical) mappa mundi represents an idea in a given discipline, 
which is slightly \quote{wrong} or \quote{mis-shappen} for the research 
mathematician's problem(s) at hand. The \quote{problem(s) at hand} are 
very unlikely to be well represented (if at all) on any existing 
mathematical mappa mundi. This is after all the \emph{point} of 
mathematical research.

\TODO{list my mappae mundi}

\subsection[title=Hilbert's program]

The primary objective of Hilbert's \quote{Program}, was to provide a 
rigorous foundation for Analysis whose metamathematics is finitist. A 
modern paraphrase of Hilbert's program would be the following 
(\emph{logicist's}) statement: 

\startBlockQuote \noindent Find a \emph{logical} axiomatization of 
mathematics whose \emph{logical} proof of consistency is 
\emph{\quote{finitist}}. \stopBlockQuote 

\noindent Hilbert was, himself, notoriously vague about what constituted a 
\quote{finitist} (logical) proof. However his program helped develop our 
current understanding of metamathematics, logical proof, as well as 
finitism. 

A statement of a \emph{computational} version of Hilbert's program is very 
similar: 

\startBlockQuote\noindent Find a \emph{computational} axiomatization of 
mathematics whose \emph{computational} proof of \emph{computability} is, 
at some meta-level, \emph{finite}. \stopBlockQuote 

A (formal) logical proof that a logical statement about a corresponding 
mathematical object, typically that the object exists and satisfies some 
collection of properties, is a textual description of a logical derivation 
of the truth of the given logical statement, using one or other particular 
logical derivation system. Using a transfinite logic, it is possible that 
the complete unfolding of the textual description might itself be 
transfinite. However the derivation that the textual description describes 
is a mathematical object in its own right and hence the truth of the 
derivation can be subject to its own proof. The distinctive structure here 
is that a mathematical object at one level is proved to exist and satisfy 
various properties, using a mathematical object (a proof derivation) at 
one meta-mathematical level higher. 

So a modern interpretation of Hilbert's Program would be to show that 
there exists a \emph{finite} derivation of the proof of the consistency 
of, for example, Zermelo–Fraenkel set theory with the Axiom of Choice 
(ZFC). The \emph{objects} in ZFC might be transfinite, and/or have 
transfinite proofs of existence, but the \emph{consistency} proof of ZFC as 
a \quote{whole} must be finite. As is well know, G\"odel's two 
Incompleteness Theorems show that any reasonably useful axiomatization of 
mathematics, has no finite proof of consistency. 

Similarly, a \emph{computational} proof that a computer program computes 
what it purports to compute is the execution of a proof verification 
program. This proof verification program, being computational can itself 
be subject to its own computational proof of correctness. Again the 
distinctive structure here is that, a program at one level is proved 
correct using a program at one meta-computational level higher. While a 
given computation might have a transfinite unfolding (or \quote{trace}), 
the description of the computation, the computations \quote{program} is a 
textual object which can be manipulated by another program. To be 
\quote{finitist}, at some meta-compuational level the proof of correctness 
must have a finite trace. 

\TODO{discuss the concept of data versus processes. classical computation 
theory has been mostly focused upon data not processes.} 

The classical Reals, which are required for classical Analysis, are 
uncountable. Once we have defined the Ordinals, we will find that there is 
a reasonable definition of $\lambda$-computation for each ordinal, 
$\lambda$. \emph{If} we assume a computational equivalent to the 
\quote{standard} Axiom of Choice, then we can define the transfinite 
ordinals and hence computational structures which can interpret the 
uncountable collection of the Reals (as \emph{data}). Using this 
structure, we will then be able to develop the classical theory of 
Analysis, which was Hilbert's ultimate goal. 

Alternatively, we can define the Reals as (measurement) \emph{processes}, 
with out requiring any \emph{transfinite} ordinals. The resulting theory 
of Analysis will not be quite classical, since the \emph{internal} logic 
of the collection of processes, is not classical. However, I conjecture 
that this process logic provides an explanation of the \quote{strangeness} 
of, and hence the correct foundations for, Quantum Mechanics. 

\TODO{Need to introduce collection of processes as a (co)algebraic 
collection/structure.... we distinguish different processes by observing 
them...} 
\TODO{For this work the focus upon the well-founded/data/algebraic versus 
the non-well-founded/process/co-algebraic is all pervasive. } 

\TODO{Categorical thought == structuralist point of view. Quote 
\cite{awodey2009a}.} 
\TODO{discuss reals as data versus reals as processes == imprecise Reals.} 

\subsection[title=Identity]

\TODO{Talk about identity of a shell/pea game. At the quark level of 
description there is no shell, pea or table to \quote{identify}. A a 
\quote{human}level, we use mental models to track our understanding of the 
effects of the shells and table upon the motion and eventual location of 
the \quote{pea}.} 

\TODO{Our usual notion of identity depends upon our intent. Do we care 
about the pea in the shell/pea game. Is there some gain for us in keeping 
track of the pea?} 

\TODO{The identity of a \quote{thing} usually does \emph{not} extend 
across too many levels of description.} 

\TODO{Mathematically same means that at some metalevel we have a finite 
collection of tests.} 

