% A ConTeXt document [master document: joyLoL.tex ]


\startcomponent joyLoLSyntax

\startchapter[title=The syntax and semantics of JoyLoL]

\startsection

The formal semantics of the \emph{other} programming languages are 
certainly simpler. This is because, the semantics of the other languages, 
are defined in terms of the semantics of JoyLoL. This means that any deep 
subtleties are simply encapsulated in the semantics of JoyLoL. Any reader 
who is willing to take the existence of the formal semantics of JoyLoL as 
given, are welcome to skip to the other chapters until they are ready or 
willing to understand the full import of the semantics of JoyLoL. 

The essential subtleties of our formal semantics for JoyLoL comes from three areas:

\startitemize[n,packed]

\head {\bf Foundations}

We are explicitly working \emph{without} classical first order set theory. 
Yet formal semantics \emph{is} defined as \emph{mappings} between 
\emph{collections} of \quote{things}. 

\noindentation \emph{How do we define mappings and collections while we 
are defining semantics with which to define mappings and collections with 
which... ?}\footnote{I personally know of no completely rigorous 
exposition of the foundations of classical mathematics. I suspect the 
closest, mathematically, we might come close to this is in G\"odel's proofs 
first and second incompleteness results. However, even here the results 
hinge upon an informal interpretation of \quote{truth} at the meta-level.} 

\head {\bf Metamathematics} 

A partial answer to our foundational question above

Any formal semantics is a programming language \emph{about} another 
\quote{object-level} programming language (which is itself a language 
\emph{about} objects as \quote{values}). 

\head {\bf Computation of \emph{properties} of potentially non-terminating 
processes} 



\stopitemize


We begin by listing the syntax of JoyLol together with its associated 
semantics. We loosely follow the presentation in 
\cite[winskel1993formalSemanticsProgrammingLanguages]. 

\stopsection

\stopchapter

\stopcomponent