% A ConTeXt document [master document: hilbertsProgram.tex]

\chapter[title=Preface] 

Despite the wealth of fruitful work inspired by Hilbert's Program, it is 
generally assumed that G\"odel's two Incompleteness Theorems show that 
Hilbert's Program has failed. The purpose of this document is to show that 
while Hilbert's \emph{logical} Program is doomed to failure, a 
\emph{computational} version of Hilbert's Program is very definitely 
capable of providing a \emph{neo-platonist} foundation of mathematics.

The important distinction here is that G\"odel's theorems show that the 
set of \quote{True} statements of a given \emph{consistent} formal theory 
which is at least as powerful as Peano Arithmetic, is a recursively 
enumerable set which is not recursive. This is what makes Hilbert's 
\emph{logical} Program fail. In this analysis, one of the two key 
requirements is that the formal theory is \emph{consistent}. That is you 
can never \quote{prove} both a statement and its negation. 

By working \emph{computationally}, we can, with an assumption equivalent 
to the Axiom of Choice, build \emph{recursive} sets which can provide 
interpretations of standard Zermelo–Fraenkel set theory (with the Axiom 
of Choice). For each ordinal, $\lambda$, we can define 
$\lambda$-computation. Standard computation is then $\omega$-computation. 
Using a computationally natural form of Vop\v{e}nka's Principle, we can 
build recursive sets of the size of a Vop\v{e}nka cardinal into which you 
can build an interpretation of Zermelo–Fraenkel set theory (ZFC), hence 
showing that ZFC is (\emph{computationally}) consistent.

By taking Cantor's Absolutely infinite multiplicities (AIMs)\footnote{See 
Cantor's letter to Dedekind dated 1899, 
\cite{vanHeijenoort1967fregeToGodel}} seriously, we can do a significant 
amount of classical analysis even with $\omega$-computational resources. 
For a given amount of computational resources, we build a dual pair of 
structures which correspond to sets and proper classes, or in terms of the 
terminology of Computer Science, data and processes. Categorically these 
dual structures are a dual Topos / Co-Topos pair. While the classical 
Reals, are located in the sets (Topos) of ZFC and hence obey classical 
logic, the $\omega$-computational Reals are processes (Co-Topos) and obey 
a weaker process logic. The important realization here is that the 
\quote{strangeness} of Quantum Mechanics comes from trying to understand 
the essentially process nature of Quantum Mechanics within the logic of 
classical sets. We assert that the natural outcome of developing the 
\quote{logic} of finite process structures \emph{is} Quantum Relativity. 
Unfortunatley, this is a topic for a subsequent document. 

