% A ConTeXt document [master document: joyLoL.tex ]

\chapter[title=CPU Model]

In \cite{gaito2018hilbertsProgramRevisitedBaseCase} we modelled the 
\emph{simple} model of computation which we call \joylol. However there 
are no implementations of this \joylol\ model of computation in a 
commercially available CPU. In this chapter we define a model of an 
idealized \emph{commercially} available CPU upon which we can rigorously 
implement pure \joylol. 

In both our models of computation, pure \joylol\ and an idealized 
commercial CPU, the important distinction is the underlying memory models. 
The pure \joylol\ model of computation is essentially a \emph{restricted 
access} Harvard model of memory. The pure \joylol\ model has a pair of 
memories one each for the data and the process. Each memory is structured 
as a stack for which only the \quote{top} of the stack can be directly 
accessed. 

We tend to assume that our commercial CPUs have a von Neumann model of 
memory consisting of \emph{one, large, randomly accessible} memory at the 
Instruction Set Architectural (ISA) interface\footnote{Note that most 
modern CPUs actually have a modified Harvard architecture at level of the 
the underlying micro-architecture, since they implement a pair of data and 
instruction caches between the Random Access Memory (RAM) and the CPU 
itself. We will, for our purposes, ignore this underlying 
micro-architecture since we are only interested in modelling the ISA 
interface. At the level of the ISA, both data and instructions are 
\quote{loaded} from \quote{one} \quote{large} (randomly accessible) Memory 
(RAM).}. As we will discuss below, for a mathematician, this assumption, 
of a \emph{single, large, randomly accessible} memory, is misguided. 

For a 64 bit CPU, the \emph{maximal} addressable memory is 16 Exabytes. 
For a mathematician, this is \emph{tiny} when compared to $\omega$. We 
could model a 128 bit CPU, or, an $x$ bit CPU for any particular value of 
$x$ we might choose, however for any \emph{finite} value of $x$, we will 
still have a \emph{maximal} addressable memory which is \emph{tiny} 
compared to $\omega$. While we could consider an idealized $\omega$ bit 
CPU, no such \emph{implementable} CPU exists. Our goal here is to model a 
\emph{realizable} CPU upon which \joylol\ might be able to run 
efficiently. 

\section[title=Model details]

\startitemize[n]

\item Storage

\startitemize[n]

\item uint64, uint32, uint16, uint8

\item int64, int32, int16, int8

\item byte, utf8Char, char

\item at the moment we do \emph{not} model floats or doubles

\item we also do \emph{not} explicitly model pointers. (Should we?)

\item arrays

\item structures

\stopitemize

\item Actions

\startitemize[n]

\item array indexing

\item addition, subtraction, mutiplication, division

\item and, or, xor, ...

\item assignments

\item functions

\stopitemize

\stopitemize

\section[title=Memory Model]

In \joylol\ the memory model, we have unlimited memory but it is only 
directly accessible at the tops of the data and process stacks. 

In \joylolRM\ the memory model, we have unlimited memory which is again 
only directly accessible at the tops of a small number of stacks which 
include \joylol's data and process stacks. The non-data, non-process 
stacks are referred to as registers or local variables.

\section[title=Call patterns]

In \joylol, \joylol\ words pass their parameters on the data or process 
stack\footnote{Context changing words are allowed to return the new 
context via a third \quote{return} parameter. However this special 
parameter could be modelled as the top of the data stack.}. 

In \joylolRM\ there are two possible call patterns, as a \joylol\ word, or 
inline. When called as a \joylol\ word, parameters can only be passed on 
the \joylol\ data and process stacks. When \quote{called inline}, the 
implementation of all callee \joylol\ words are embedded inline into the 
body of the implementation of the caller \joylol\ word. 

\TODO{What about more traditional C-like call patterns? Can we interpret 
them as implicitly pushing the arguments on to the data stack? \emph{Most} 
words simply manipulate the top few items on the data stack, and return 
\emph{one} result. This is essentially the standard C-like call pattern. 
Lua allows multiple return values:}

\starttyping
x, y, z = aCall(a, b, c)
\stoptyping

\TODO{This leaves one case: how do we denote the \emph{infinite} change to 
the data stack. Effectively this can only be the removal of the whole 
stack, AKA clearData.} 

\TODO{What about words that manipulate the process stack? I think most 
such words actually always manipulate the data stack as well. How are 
these represented in a C-like way? Since the process stack \emph{is} the 
continuation structure of a method, the Lua multi-return is not 
appropriate for the process changes. However the multi-return does suggest 
a way forward: the multi-return is all about what will be done with the 
results on the data stack. In process terms this (might) translate into 
control structures. However a problem here is that the callee word has the 
arbitrary ability to \emph{change} the contents of the process stack.. and 
so what happens once the callee returns. This is difficult if we equate 
the contents of the process stack as a control structure compiled 
\emph{before} the callee word gets called. Is it this difficulty which 
precludes some words being compiled? (see below)} 

\TODO{When does a word get too complex to be compiled (inline)?}

